---
title: "ASM -  Local Poisson regression"
author: "Sergi Carol, Laura Cebollero, Alex Rodriguez"
date: "10th November, 2018"
output: 
    pdf_document:
        number_sections: yes
---

```{r setup, include=FALSE}
rm(list = ls())

# Load and install necessary packages
requiredPackages <- c("sm", "readr")

for (pac in requiredPackages) {
    if(!require(pac,  character.only=TRUE)){
        install.packages(pac, repos="http://cran.rstudio.com")
        library(pac,  character.only=TRUE)
    } 
}
rm(pac)
rm(requiredPackages)
```


# Introduction

In this delivery we are asked to study local Poisson Regression. More specifically, the aim of this report is two-fold:

- Create a function to choose the best bandwidth for a Poisson.
- Use said function on the Country Development dataset.

# Bandwidth choice for local Poisson Regression

We have modified the already seen functions so that they can be used with data following a Poisson distribution.

To do so, we have had to change the leave-one-out CV estimation of the expected log-likelihood following the formula given in the delivery statement. The resulting code of such function is:

```{r}
loglik.CV <- function(x,y,h){
    n <- length(x)
    pred <- sapply(1:n, 
                   function(i,x,y,h){
                       sm.poisson(x=x[-i],y=y[-i],h=h,eval.points=x[i],display="none")$estimate
                   },   x,y,h)
    total_sum = 0
    for (i in 1:n){
      total_sum = total_sum + (log((exp(-pred[i]) * ((pred[i]**y[i])/factorial(y[i])))^-i))/n
    }
    return(total_sum)
}
```

Now, since we have the log-likelihood method for the Poisson distribution, we can compute the best bandwidth using it:
```{r}
h.cv.sm.poisson <- function(x,y,rg.h=NULL,l.h=10,method=loglik.CV){
  cv.h <- numeric(l.h)
  if (is.null(rg.h)){
    hh <- c(h.select(x,y,method="cv"),
            h.select(x,y,method="aicc"))#,hcv(x,y))
    rg.h <- range(hh)*c(1/1.1, 1.5)
  }
  i <- 0
  gr.h <- exp( seq(log(rg.h[1]), log(rg.h[2]), l=l.h))
  for (h in gr.h){
    i <- i+1
    cv.h[i] <- method(x,y,h)
  }
  return(list(h = gr.h, 
              cv.h = cv.h, 
              h.cv = gr.h[which.min(cv.h)]))
}
```

Now that we have them implemented we can proceed to use them for data following a Poisson.

# Local Poisson regression for Country Development data

