---
title: "ASM -  Local Poisson regression"
author: "Sergi Carol, Laura Cebollero, Alex Rodriguez"
date: "10th November, 2018"
output: 
    pdf_document:
        number_sections: yes
---

```{r setup, include=FALSE}
# Load and install necessary packages
requiredPackages <- c("sm", "readr")

for (pac in requiredPackages) {
    if(!require(pac,  character.only=TRUE)){
        install.packages(pac, repos="http://cran.rstudio.com")
        library(pac,  character.only=TRUE)
    } 
}
rm(pac)
rm(requiredPackages)

library("knitr")
```


# Introduction

In this delivery we are asked to study local Poisson Regression. More specifically, the aim of this report is two-fold:

- Create a function to choose the best bandwidth for a Poisson.
- Use said function on the Country Development dataset.

# Bandwidth choice for local Poisson Regression

We have modified the already seen functions so that they can be used with data following a Poisson distribution.

To do so, we have had to change the leave-one-out CV estimation of the expected log-likelihood following the formula given in the delivery statement. The resulting code of such function is:

```{r}
loglik.CV <- function(x,y,h){
    n <- length(x)
    pred <- sapply(1:n, 
                   function(i,x,y,h){
                       sm.poisson(x=x[-i],y=y[-i],h=h,eval.points=x[i],display="none")$estimate
                   },   x,y,h)
    total_sum = 0
    for (i in 1:n){
      total_sum = total_sum + (log((exp(pred[i]) * ((pred[i]**y[i])/factorial(y[i])))))/n
      # Incorrect:
      # total_sum = total_sum + (log((exp(-pred[i]) * ((pred[i]**y[i])/factorial(y[i])))^-i))/n
    }
    return(total_sum)
}
```

Now, since we have the log-likelihood method for the Poisson distribution, we can compute the best bandwidth using it:
```{r}
h.cv.sm.poisson <- function(x,y,rg.h=NULL,l.h=10,method=loglik.CV){
  cv.h <- numeric(l.h)
  if (is.null(rg.h)){
    hh <- c(h.select(x,y,method="cv"),
            h.select(x,y,method="aicc"))#,hcv(x,y))
    rg.h <- range(hh)*c(1/1.1, 1.5)
  }
  i <- 0
  gr.h <- exp( seq(log(rg.h[1]), log(rg.h[2]), l=l.h))
  for (h in gr.h){
    i <- i+1
    cv.h[i] <- method(x,y,h)
  }
  return(list(h = gr.h, 
              cv.h = cv.h, 
              h.cv = gr.h[which.min(cv.h)]))
}
```

Now that we have them implemented we can proceed to use them for data following a Poisson.

# Local Poisson regression for Country Development data

Let's first load the countries data:
```{r warning=F}
countries <- na.omit(read_table2("countries.csv"))
summary(countries)
head(countries)
attach(countries)
```

Since it contains some missing values we are going to omit them on their reading.

We are asked to predict the `le.fm` variable. However, looking at the summary we can see that it ranges from -1 to 11. Thus, meaning that it contains a negative value.

So a new variable, as suggested is defined such that
```{r}
le.fm.0 <- pmax(0,le.fm)
```

What `pmax` does is return the parallel maxima and minima of the input values, which in this case it is `le.fm`. So we are forcing the minimum value -1 to be a 0, leaving the maximum as it was:

```{r}
cat("MAXIMUM le was ", max(le.fm)," and now still is", max(le.fm.0) , "\n")

cat("MINIMUM le was ", min(le.fm)," and now is", min(le.fm.0) , "\n")
```

Now let's proceed onto fitting a local Poisson regression using our functions onto our data.

More specifically, we are going to fit a local Poisson regression modelling our new variable `le.fm.0` as a function of `life.exp`. 

This is the data we want to model:
```{r}
plot(life.exp, le.fm.0)
```

Let's first choose the best bandwidth. To preserve locality, we establish possible h that range from 2 to 15.

```{r}
range.h = c(2, 15)
h.CV.loglik <- h.cv.sm.poisson(life.exp,le.fm.0 ,rg.h=range.h, method=loglik.CV)
```

```{r}
plot(h.CV.loglik$h,h.CV.loglik$cv.h)
lines(h.CV.loglik$h,h.CV.loglik$cv.h)
```

```{r}
kable(data.frame(h.CV.loglik$h, h.CV.loglik$cv.h))
```

We can see how the minimum log-likelihood is found close to 3, so we are going to use that value.

Let's see how our Poisson regression would be using that bandwith:
```{r}
chosen.h =  h.CV.loglik$h.cv
m1 <- sm.poisson(life.exp, le.fm.0, h=chosen.h)
```

We can see from the plot above how the line fits our data without the line being not smooth. So the optimal bandwith is:

```{r}
chosen.h
```

We can see how the regression does not go below 0, as we wanted.