---
title: "ASM - Ridge Regression"
author: "Sergi Carol, Laura Cebollero, Alex Rodriguez"
date: "8th November, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
source("parameter_lambda.R")
```

## Some notes 

All the code related to this report and delivery can be found on the files:

- `parameter_lambda.R`. Has the code for the 1st task: choosing the $\lambda$ penalization parameter.
- `all_functions.R`. Contains all the abstracted functions used to compute the MSPE with different methods.
- `boston.R`. Has the code related to the 2nd task: the Boston Housing dataset Ridge Regression study. Uses the functions on `all_functions.R`.
- `ASM_report_RR.Rmd`. The RMarkdown that has produced this .pdf formatted report.



# Choosing the penalization parameter

As asked, we have written two functions to choose an adequate parameter $\lambda$ 
for the Ridge Regression.

The implementation of such functions can be found at `parameter_lambda.R` with
its corresponent comments inline to understand how the code works.

We use 25 different $\lambda$ that range from 0 to 100.000.

## Choosing $\lambda$ with MPSE and a validations set
The first function implemented is the one that allows us to choose $\lambda$ using a validation set.

Such validation set has been selected by using the column `train`, which tells us which
data can be used for training purposes.

This leaves us with a validation set of 30 observations, and 67 observations as the
training sample.

Below is the plotting the resulting MSPE values for each lambda:

```{r plot_val, echo=FALSE, out.width="250px", out.height="250px", fig.align="center"}
plot(log(1+lambda.v), MSPE.val)
abline(v=log(1+lambda.val),col=2,lty=2)
```
```{r min}
(minPos = which.min(MSPE.val))
lambda.v[minPos]
```
We can see that using this method, the $\lambda$ that gives us the least error is 
the 8th one, with a value of 27.7.

## Choosing $\lambda$ with MPSE and K-fold cross-validation

We have also done the implementation of k-fold cross-validation. In this case we split the dataset in K differnet parts, then we train the model using K - 1 parts and leave the remaining one to validate the model, thats it, calculating the MSE. Once we have calculated the value of the $\lambda$ for each fold, we choose the value $\lambda$ from each iteration of k that gives the less error.

Using a 10-fold CV, we obtain the following plot:

```{r plotcv10, echo=FALSE, out.width="250px", out.height="250px", fig.align="center"}
plot(log(1+lambda.v), mspe.10)
abline(v=log(1+lambda.mspe.10),col=2,lty=2)
```

```{r}
(minPos = which.min(mspe.10))
lambda.v[minPos]
```

It seems that in this case, we are selecting the one with $\lambda = 0$, which means
we are not penalizing.


If we use a 5-fold CV:

```{r plotcv5, echo=FALSE, out.width="250px", out.height="250px", fig.align="center"}
plot(log(1+lambda.v), mspe.5)
abline(v=log(1+lambda.mspe.5),col=2,lty=2)
```

```{r}
(minPos = which.min(mspe.5))
lambda.v[minPos]
```
Now the selected  $\lambda$ with the least error is the one on the 9th position with
 $\lambda = 45.41$.
 
 Let's now proceed onto compare those results with those obtained from using the methods of leave-one-out cross-validation and Generalized Cross-Validation.
 
```{r}
mspe.cv <- MSPEcv(X, Y, n.lambdas, lambda.v, n)
mspe.gcv <- MSPEgcv(X, Y, n.lambdas, lambda.v, beta.path, diag.H.lambda, n)
```

```{r plotting2, echo=FALSE,  out.width="250px", out.height="250px", fig.align="center"}
plot(df.v, MSPE.val)
points(df.v, mspe.cv,col=13,pch=19,cex=.75)
points(df.v, mspe.5,col=12,pch=19,cex=.75)
points(df.v, mspe.10,col=8,pch=19,cex=.75)
points(df.v, mspe.gcv,col=10,pch=19,cex=.5)

legend("topright",
       c("MSPE.val","MSPE.CV", "MSPE 5-Fold", "MSPE 10-Fold", "MSPE GCV"),
       pch=c(19, 19, 19, 19, 19),
       lty=c(0,0,0,0,0),
       lwd=c(0,0,0,0,0),
       col=c(9, 13, 12, 8, 10)
       )

```


TODO: Interpret plot


\newpage

# Ridge Regression for Boston Housing Area
 Now, we are going to proceed to apply Ridge Regression to the Boston Housing Dataset.
 
 **Please note that we have not found uploaded on `El Racó` the corrected dataset.**
 Thus, we have **worked using the dataset provided by the `MASS` library**.

To apply the Ridge Regression method onto this dataset, we have created a matrix $X$ which 
contains all the 13 explanatory variables. This matrix has been centered and scaled.

We have then created a vector $Y$ which is the response variable. It has been centered but not scaled.

```{r include=FALSE, echo=FALSE}
source("boston.R")
```

```{r plotting, echo=FALSE,  out.width="300px", out.height="300px", fig.align='center'}
plot(df.v, MSPE.val)
points(df.v, mspe.cv,col=13,pch=19,cex=.75)
points(df.v, mspe.5,col=12,pch=19,cex=.75)
points(df.v, mspe.10,col=8,pch=19,cex=.75)
points(df.v, mspe.gcv,col=10,pch=19,cex=0.5)

abline(v=df.GCV,col=1,lty=2,lwd=2)
abline(v=0.001,col=2,lty=2,lwd=1, cex=0.35) # Since df.CV.H.lambda == 0, this is to force the print of this abline
# abline(v=df.CV.H.lambda + 0.001,col=2,lty=2, lwd=3)

legend("topright",
       c("MSPE.val","MSPE.CV", "MSPE 5-Fold", "MSPE 10-Fold", "MSPE GCV", "lambda.GCV","lambda.CV"),
       pch=c(1, 19, 19, 19, 19, NA, NA),
       lty=c(0,0,0,0,0,2,2),
       lwd=c(0,0,0,0,0,2,1),
       col=c(9, 13, 12, 8, 10, 1, 2)
       )

```

 We can see that we get higher values on the  MSPE 5-Fold, the lowest ones obtained from
 MSPE val (that using, using an 80% of the data as training and 20% of the original dataset as a validation set).
 
In this case, the 10-fold method does not prove as a good method for it declines until reaching 0.

The generalized Cross-validation method as well as the leave one out (MSPE.CV) follow close
the MSPE 5-fold evolution and in fact have a greater value when df.v > 9.