---
title: "ASM - Ridge Regression"
author: "Sergi Carol, Laura Cebollero, Alex Rodriguez"
date: "8th November, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
source("parameter_lambda.R")
```

## Some notes 
All the code related to this report and delivery can be found on the files:

- `parameter_lambda.R`. Has the code for the 1st task: choosing the $\lambda$ penalization parameter.
- `all_functions.R`. Contains all the abstracted functions used to compute the MSPE with different methods.
- `boston.R`. Has the code related to the 2nd task: the Boston Housing dataset Ridge Regression study. Uses the functions on `all_functions.R`.
- `ASM_report_RR.Rmd`. The RMarkdown that has produced this .pdf formatted report.



# Choosing the penalization parameter

As asked, we have written two functions to choose an adequate parameter $\lambda$ 
for the Ridge Regression.

The implementation of such functions can be found at `parameter_lambda.R` with
its corresponent comments inline to understand how the code works.

We use 25 different $\lambda$ that range from 0 to 100.000.

## Choosing $\lambda$ with MPSE and a validations set
The first function implemented is the one that allows us to choose $\lambda$ using a validation set.

Such validation set has been selected by using the column `train`, which tells us which
data can be used for training purposes.

This leaves us with a validation set of 30 observations, and 67 observations as the
training sample.

Below is the plotting the resulting MSPE values for each lambda:

```{r plot_val, echo=FALSE}
plot(log(1+lambda.v), MSPE.val)
abline(v=log(1+lambda.val),col=2,lty=2)
```
```{r min}
(minPos = which.min(MSPE.val))
lambda.v[minPos]
```
We can see that using this method, the $\lambda$ that gives us the least error is 
the 8th one, with a value of 27.7.

## Choosing $\lambda$ with MPSE and K-fold cross-validation

We have also done the implementation of k-fold cross-validation. 

Using a 10-fold CV, we obtain the following plot:

```{r plotcv10, echo=FALSE}
plot(log(1+lambda.v), mspe.10)
abline(v=log(1+lambda.mspe.10),col=2,lty=2)
```

```{r}
(minPos = which.min(mspe.10))
lambda.v[minPos]
```

It seems that in this case, we are selecting the one with $\lambda = 0$, which means
we are not penalizing.


If we use a 5-fold CV:

```{r plotcv5, echo=FALSE}
plot(log(1+lambda.v), mspe.5)
abline(v=log(1+lambda.mspe.5),col=2,lty=2)
```

```{r}
(minPos = which.min(mspe.5))
lambda.v[minPos]
```
Now the selected  $\lambda$ with the least error is the one on the 9th position with
 $\lambda = 45.41$.
 
 
 TODO: Compare results with those obtained with leave 1 out and CV.
 
 
 
 # Ridge Regression for Boston Housing Area
 Now, we are going to proceed to apply Ridge Regression to the Boston Housing Dataset.
 
 **Please note that we have not found uploaded on `El Racó` the corrected dataset.**
 Thus, we have **worked using the dataset provided by the `MASS` library**.

To apply the Ridge Regression method onto this dataset, we have created a matrix $X$ which 
contains all the 13 explanatory variables. This matrix has been centered and scaled.

We have then created a vector $Y$ which is the response variable. It has been centered but not scaled.




 