---
title: "GAM Model"
author: "Sergi Carol Laura Cebollero Alex Rodriguez"
date: "December 30, 2018"
output: pdf_document
---

```{r setup, include=FALSE}
library(mgcv)

set.seed(123)
knitr::opts_chunk$set(echo = TRUE)
```

## Introduction
The aim of this lab is to understand the GAM model and get an insight into the fitted model. The lab will consist of two different tasks, the first one will be to create different GAM models and see how well they fit the model. The second one will be to choose the best model according to our criteria.

### Exercise 1

First of all, we are going to read the hirsutism dataset and explore it:

```{r}
hirs <- read.table("hirsutism.dat",header=T, sep="\t",fill=TRUE)
hirs$Treatment <- as.factor(hirs$Treatment)

summary(hirs)
head(hirs)
attach(hirs)

boxplot(hirs[,2:5])
```

We can see from the summary and the boxplots that the FG at the start of the treatment (FGm0) is overall  greater than afterwards, where the patient has received the treatment.

In fact, we can see overall how there is a decrease of the FG as the treatment has an effect on the patient's hirsutism.
```{r}

par(mfrow=c(2,2))
boxplot(hirs[,2]~Treatment,ylim=c(0,30), main="FG m0")
boxplot(hirs[,3]~Treatment,ylim=c(0,30), main="FG m3")
boxplot(hirs[,4]~Treatment,ylim=c(0,30), main="FG m6")
boxplot(hirs[,5]~Treatment,ylim=c(0,30), main="FG m12")
par(mfrow=c(1,1))

par(mfrow=c(2,2))
boxplot(hirs[Treatment==0,2:5],ylim=c(0,30), main="Treatment 0")
boxplot(hirs[Treatment==1,2:5],ylim=c(0,30), main="Treatment 1")
boxplot(hirs[Treatment==2,2:5],ylim=c(0,30), main="Treatment 2")
boxplot(hirs[Treatment==3,2:5],ylim=c(0,30), main="Treatment 3")
par(mfrow=c(1,1))
```

Now if we take a look at the boxplots above where we are comparing the FG at different stages (months 0, 3, 6 and 12)
with four different treatments.

We can see how the tren of diminishing hirsutism may vary depending on the treatment the patient receives.
For example, with treatment 0 the hirsutism seems to stabilize more than on the other 3 treatments.

On treatment 1, the variance seems way greater than on the other treatments. Treatment 2 seems to not vary a lot and treatment 3 seems to have the most decrease on the hirsutism.

We want to create a model that explains the FGm12 in function of the variables that were mesured
at the beginning of the trial.

### First model
Let's start with a simple model.
```{r}
am1.0 <- gam(FGm12 ~ weight + height + DiaPres + SysPres + FGm0 + Treatment, data=hirs)
am1.0
summary(am1.0)
```
We can see how the weight, height, diaPres and SysPres are not relevant, and are only FGm0 and the treatments.
So let's try to achieve a simpler model without the irrelevant variables.

```{r}
am1.1 <- gam(FGm12 ~FGm0 + Treatment, data=hirs)
am1.1
summary(am1.1)

```
This model does indeed look better and R squared is  greater than before.

Now let's try yet another model with a smooth GAM with FGm0 for each treatment:
```{r}
am1.2 <- gam(FGm12 ~ s(FGm0, by=Treatment) + Treatment, data=hirs)
am1.2
summary(am1.2)
plot.gam(am1.2, page=1, residuals=TRUE, shade=TRUE) 
vis.gam(am1.2)
```
This model is quite interesting. We can see that Treatment 0 is the only one that does not seem to be linear whereas the other 3 are.


Now we are going to create yet another model with a tensor product smooth for height and weight.
```{r}
am1.3 <- gam(FGm12 ~ s(FGm0, by=Treatment) + te(weight, height), data=hirs)
am1.3
summary(am1.3)
plot.gam(am1.3, page=1, residuals=TRUE, shade=TRUE) 
```

Once again it appears that the weight and height are not relevant.

Finally, we create the last model, with  a smooth for FGm0, weight and height separately.

```{r}
am1.4 <- gam(FGm12 ~ s(FGm0, by=Treatment) + s(weight) + s(height), data=hirs)
am1.4
summary(am1.4)
plot.gam(am1.4, page=1, residuals=TRUE, shade=TRUE) 
```
Yet again, height and weight do not seem that relevant by looking at the p-value. However the R squared adjusted has increased again, which means the proportion of variance explained by this model is greater than before and around almost 37%.


## Task 2

Now we are going to use ANOVA to select the best model that best fits our data at hand and explains the hirsutism at the 12th month.