---
title: "Homework 1"
author: "Laura Cebollero Ruiz, Alexandre Rodríguez Garau"
geometry: "left=3cm,right=3cm,top=2cm,bottom=2cm"
date: "15/10/2018"
output: pdf_document
---



```{r setup, include=FALSE}
requiredPackages <- c("ggplot2", "car", "MASS", "devtools", "readr", "knitr", "easyGgplot2", "gridExtra", "doBy")

for (pac in requiredPackages) {
    if(!require(pac,  character.only=TRUE)){
        install.packages(pac, repos="http://cran.rstudio.com")
        library(pac,  character.only=TRUE)
    } 
}
rm(pac)
rm(requiredPackages)


# Set WD and load data
wd = getwd()
if(grepl("nora", wd)) {
    setwd("~/Documents/18-19/ASM/HW/01")
} else {
  #setwd("")
}
rm(wd)
```

# Part A

First of all we will load the data. As described by  the problem statement, the explanatory
variables are `dia` and `Ferm` and the response vars are `pH`, `strep` and `lactob`.

Since we know beforehand that the `Ferm` variable, which refers to
Fermentation temperature, describes on to which group the Iogurt is from, we interpret it as a factor (a.k.a level).

The same interpretation rule is applied to dia, for it is not a continuous variable
(a iogurt cannot be fermented by 3.5 days, for example).

```{r message=FALSE}
Iogurt <- read_delim("Iogurt.csv", ";", escape_double = FALSE, 
    trim_ws = TRUE)
Iogurt$Ferm = as.factor(Iogurt$Ferm)
Iogurt$dia = as.factor(Iogurt$dia)
```

After deciding the type of each variable we can proceed onto taking a look at its 
summary.
```{r }
kable(summary(Iogurt))
```

For the `Ferm` we can see that there are 2 groups and each has the same size of 30 individuals per group.

We can see that the mean of the variable `pH` is 378 but the minimum value is 4, and its maximum is 448.
For the variable `strep`, with a mean of 671 its minimum is 71 and its maximum 782.
And finally, for the variable `lactob`, its mean is 719 with a minimum of 74 and a maximum of 788. 

This could indicate that we might have some **outliers** in our dataset for both the mean
and median are **skewed and biased towards the maximum of each one of these 3 variables.**


## Visual analysis
In order to see it in a visual way we will plot the boxplots:

```{r}
par(mfrow=c(1,2))

boxplot(Iogurt$pH ~ Iogurt$Ferm, 
        main="pH ~ Fermentation Temp.", 
        xlab="Fermentation Temperature", 
        ylab="pH")
boxplot(Iogurt$lactob ~ Iogurt$Ferm, 
        main="Lactobacillus ~ Fermentation Temp.", 
        xlab="Fermentation Temperature",
        ylab="Lactobacillus concentration")
boxplot(Iogurt$strep ~ Iogurt$Ferm, 
         main="Streptococus ~ Fermentation Temp.", 
        xlab="Fermentation Temperature",
        ylab="Streptococus concentration")
```


We can see clearly that there are some values that really differ from the rest of the values by many standard deviations. Hence we can safely delete these values for we deem them as outliers and they bias our data, thus probably biasing our possible future modeling of it.

So we have put a threshold on 200 units, but it could be higher, like 300. We are using it just to
disregard the lowest values.

```{r}
Iogurt <- Iogurt[Iogurt$pH     > 200, ]
Iogurt <- Iogurt[Iogurt$strep  > 200, ]
Iogurt <- Iogurt[Iogurt$lactob > 200, ]
```

```{r}
par(mfrow=c(1,2))

boxplot(Iogurt$pH ~ Iogurt$Ferm, 
        main="pH ~ Fermentation Temp.", 
        xlab="Fermentation Temperature", 
        ylab="pH")
boxplot(Iogurt$lactob ~ Iogurt$Ferm, 
        main="Lactobacillus ~ Fermentation Temp.", 
        xlab="Fermentation Temperature",
        ylab="Lactobacillus concentration")
boxplot(Iogurt$strep ~ Iogurt$Ferm, 
         main="Streptococus ~ Fermentation Temp.", 
        xlab="Fermentation Temperature",
        ylab="Streptococus concentration")
```

Now that we have removed the outliers, let's take a look at the plot of the different response variables with the explanatory variable `dia` in the x axis . Also, we will use the `Ferm` variable to see which individials belong to one group or the other. 

```{r  out.width="300px", out.height="300px", fig.align="center"}
ggplot(Iogurt, aes(x=dia, y=pH, color=Ferm)) + 
    geom_point(shape=factor(Iogurt$Ferm), size = 2, stroke = 1.3) +
    ggtitle("pH ~ dia")+
    scale_colour_hue(l=50) + 
    geom_smooth(method=lm,   
                se=FALSE,    
                fullrange=TRUE) 

```
From this plot we can see that for dia <= 7, the iogurts on the fermentation group 
with temperature 42 have always a greater pH than those with a fermentation temperature
of 43.5 Celsius degrees .

But while the dia increases, the pH on both groups clearly and steadily decreases 
and at the 2nd week they are not separated anymore, for we can see that the two fermentation
groups are mixed on the pH axis.
```{r  out.width="300px", out.height="300px", fig.align="center"}
ggplot(Iogurt, aes(x=dia, y=strep, color=Ferm)) + 
    geom_point(shape=factor(Iogurt$Ferm), size = 2, stroke = 1.3) +
    ggtitle("strep ~ dia")+
    scale_colour_hue(l=50) + 
    geom_smooth(method=lm,   
                se=FALSE,    
                fullrange=TRUE)
```
When comparing the strep bacteria presence on each week, we can see that
its presence is higher on the first day but it kind of stabilizes on the 7th day
and decreases slightly over every week, so if we imagine a visual slope, it would not be
a slope with a great value.

On the first day, generally speaking the strep is higher when the fermentation 
temperature is lower.

After the 7th day, the  strep slightly decreases as we have said, but more importantly
the variance on those iogurts with a fermentation temperature of 42 is way greater 
than the one of temperature 43.5, since they are more scattered along the Y axis,
ranging from a ~740 units of strep to 700, whereas the range of the group of temperature
43.5 ranges from ~735 to ~710.

```{r  out.width="300px", out.height="300px", fig.align="center"}
ggplot(Iogurt, aes(x=dia, y=lactob, color=Ferm)) + 
    geom_point(shape=factor(Iogurt$Ferm), size = 2, stroke = 1.3) +
    ggtitle("lactob ~ dia")+
    scale_colour_hue(l=50) + 
    geom_smooth(method=lm,   
                se=FALSE,    
                fullrange=TRUE) 
```
Finally, in the case of lactob presence,
we can see that the 2 groups are clearly distinguished on the zeroeth day, with its presence being
way greater on the 42Temperature group, arund ~785, than on the group of 43.5 where they are scattered
around ~770 units.

## Table analysis
Now we will create a table where we will see the basic descriptive measures for every possible value of day.

If we take a look at the variable dia we can see that it ranges from 0 to 28 but only takes the values 0, 7, 14, 21 and 28, reason for
why we have chosen at the start to deem it as a factor.

```{r}
Idf <- as.data.frame(Iogurt)
kable(te <- summaryBy(cbind(pH, lactob, strep)~ dia * Ferm, data = Idf, FUN = c(mean, sd)))
```

As we can see in the table we have 10 rows. Each row shows the mean and the standard deviation of the observations from a determined day ---0, 7, 14, 21 or 28--- and type of fermentation ---T42 or T43.5, which are our response variables.

Since `dia` has 5 possible values and `Ferm` has two, we obtain $2\cdot5 = 10$ possible combinations.

We will focus on each dia and then for each group inside that dia.

### Dia 0
- Mean on pH is greater for T42.
- Mean on lactob is greater for T42.
- Mean on strep is greater for T42.

- **pH std. dev is greater for T43.5, so it varies more on T43.5.**
- lactob std. dev is greater for T42.
- strep std. dev is greater for T42.

## Dia 7
- Mean on pH is greater for T42.
- Mean on lactob is greater for T42.
- Mean on strep is greater for T42.

- **pH std. dev is greater for T43.5, so it varies more on T43.5.**
- lactob std. dev is greater for T42.
- **strep std. dev is greater for T43.5**.


## Dia 14
- Mean on pH is greater for T42, but they are **very** close.
- Mean on lactob is greater for T42, but again, they are very close.
- Mean on strep is greater for T42.

- **pH std. dev is greater for T43.5, so it varies more on T43.5.**
- lactob std. dev is greater for T42.
- strep std. dev is greater for T42.

## Dia 21
- **Mean on pH is greater for T43.5, so it has reversed from other days**.
- **Mean on lactob is greater for T43.5, but only for 0.2. So we can deem them almost the same.**
- **Mean on strep is greater for T43.5, so it has reversed**.

- pH std. dev is greater for T43.5, as before.
- **lactob std. dev is greater for T43.5, so it has reversed.**
- strep std. dev is greater for T42, as before.


## Dia 28
- Mean on pH is greater for T42.
- Mean on lactob is greater for T42.
- Mean on strep  is almost the same.

- **pH std. dev is greater for T42, so it has reversed.**
- lactob std. dev is greater for T42.
- strep std. dev is greater for T42.



# Part B

Since we can assume that the variable `pH` follows a normal distribution we will proceed to check whether differences exist between the mean of the `pH` values of one group and another. In order to be sure that the variable `dia` doesn't affect us we will take all those values that belong to the day 7. 

Now we will separate the values into 2 separate groups depending on their `Ferm` value and then take only the rows that belong to day 7.


```{r}
I42 <- Iogurt[Iogurt$Ferm == "T42" ,]
I43 <- Iogurt[Iogurt$Ferm == "T43.5", ]

I42_7 <- I42[I42$dia == 7, ]
I43_7 <- I43[I43$dia == 7, ]
```

Once we have separated our data and only taken the observations that belong to day 7, we are ready to perform both T-Test and a F Test to see it there really exist any significant differences between the means and to check if the variances are equal.

```{r}
t.test(I42_7$pH, I43_7$pH)

var.test(I42_7$pH, I43_7$pH)
```

For the means we can see that the p-value that we obtain is lower than 0.05, which means that we have to reject the null hypothesis and accept that "true difference in means is not equal to 0", in other words: the means are not the same for both groups. 

However the F test does return a high p-value of 0.587 which leads to believe that the variances of both groups are indeed equal ---i.e. true ratio of variances is equal to 1. 

To move on we will check the same we just did for the other two response variables `lactob` and `strep`. Also we will check it for the rest of days.

```{r}
I42_0 <- I42[I42$dia == 0, ]
I43_0 <- I43[I43$dia == 0, ]

I42_14 <- I42[I42$dia == 14, ]
I43_14 <- I43[I43$dia == 14, ]

I42_21 <- I42[I42$dia == 21, ]
I43_21 <- I43[I43$dia == 21, ]

I42_28 <- I42[I42$dia == 28, ]
I43_28 <- I43[I43$dia == 28, ]


t.test(I42_0$lactob, I43_0$lactob)
var.test(I42_0$lactob, I43_0$lactob)

t.test(I42_14$lactob, I43_14$lactob)
var.test(I42_14$lactob, I43_14$lactob)

t.test(I42_21$lactob, I43_21$lactob)
var.test(I42_21$lactob, I43_21$lactob)

t.test(I42_28$lactob, I43_28$lactob)
var.test(I42_28$lactob, I43_28$lactob)

t.test(I42_0$strep, I43_0$strep)
var.test(I42_0$strep, I43_0$strep)

t.test(I42_14$strep, I43_14$strep)
var.test(I42_14$strep, I43_14$strep)

t.test(I42_21$strep, I43_21$strep)
var.test(I42_21$strep, I43_21$strep)

t.test(I42_28$strep, I43_28$strep)
var.test(I42_28$strep, I43_28$strep)
```

As we can see, contrary to the previous t-test, all means are equal except for the day 0 and the variable `strep`, which has a p-value of 0.0387. Also, for both `strep` and `lactob` the ratios of variances are equal to 1.

# Part C

Let's try to predict the value of the variables `strep` and `lactob` from the variable `pH`. To do that we will create a linear model for both. This linear model will look like: $$strep = \beta_1 \cdot pH + \beta_0 + e$$

```{r}
model_strep <- lm(strep ~ pH, Iogurt)
summary(model_strep)
```

According to the linear model the expression that best predicts the value of `strep` in term of `pH` is: 
$$strep = 1.244 \cdot pH + 219.99 $$
If we look at the adjusted $R^2$ we can see that the data points fit quite well the regression line. We know that if the pH of a yoghurt is lower than 4 that yoghurt is expired. If we wanted to know what the minimum value that `strep` has to have to consider a yoghurt expired we just need to substitute `pH` in the formula for 4. 

$$strep = 1.244 \cdot 4 + 219.99 = 224.966$$

According to this linear model, all those observations that have a `strep` value lower than 224.966 can be considered expired yoghurts. 

In the same way we can do the same to predict the value of `lactob` from the variable `pH`.

```{r}
model_lactob <- lm(lactob ~ pH, Iogurt)
summary(model_lactob)
```
This linear model looks like the following:

$$lactob = 0.908 \cdot pH + 375.56$$

If we do the same as before to calculate the minimum `lactob` value for an expired yoghurt we just need to plug 4 into `pH`. 

$$lactob = 0.908 \cdot 4 + 375.56 = 379.19$$

## Most accurate model

In our opinion, the most accurate model is the one predicting the lactob value, for its adjusted $R^2$ is greater with a value of  0.8844 than the one pedicting the strep variable, which has a adjusted $R^2$  of 0.862.

# Conclusions
